# Control Structures

```{r,echo=FALSE}
knitr::opts_chunk$set(comment = NA, prompt = TRUE, collapse = TRUE)
```

[Watch a video of this section](https://youtu.be/BPNLjUDZ8_o)

Control structures in R allow you to control the flow of execution of
a series of R expressions. Basically, control structures allow you to
put some "logic" into your R code, rather than just always executing
the same R code every time. Control structures allow you to respond to
inputs or to features of the data and execute different R expressions
accordingly.

Commonly used control structures are

- `if` and `else`: testing a condition and acting on it

- `for`: execute a loop a fixed number of times 

- `while`: execute a loop _while_ a condition is true 

- `repeat`: execute an infinite loop (must `break` out of it to stop)

- `break`: break the execution of a loop

- `next`: skip an iteration of a loop

Most control structures are not used in interactive sessions, but
rather when writing functions or longer expressions. However, these
constructs do not have to be used in functions and it's a good idea to
become familiar with them before we delve into functions.


## `if`-`else`

[Watch a video of this section](https://youtu.be/ZaBtJPYYGwg)

The `if`-`else` combination is probably the most commonly used control
structure in R (or perhaps any language). This structure allows you to
test a condition and act on it depending on whether it's true or
false. 

For starters, you can just use the `if` statement.

```r
if(<condition>) {
        ## do something
} 
## Continue with rest of code
```

The above code does nothing if the condition is false. If you have an
action you want to execute when the condition is false, then you need
an `else` clause.

```r
if(<condition>) {
        ## do something
} 
else {
        ## do something else
}
```

You can have a series of tests by following the initial `if` with any
number of `else if`s.

```r
if(<condition1>) {
        ## do something
} else if(<condition2>)  {
        ## do something different
} else {
        ## do something different
}
```

Here is an example of a valid if/else structure.

```{r,prompt=FALSE}
## Generate a uniform random number
x <- runif(1, 0, 10)  
if(x > 3) {
        y <- 10
} else {
        y <- 0
}
```

The value of `y` is set depending on whether `x > 3` or not. This
expression can also be written a different, but equivalent, way in R.

```{r,prompt=FALSE}
y <- if(x > 3) {
        10
} else { 
        0
}
```

Neither way of writing this expression is more correct than the
other. Which one you use will depend on your preference and perhaps
those of the team you may be working with.

Of course, the `else` clause is not necessary. You could have a series
of if clauses that always get executed if their respective conditions
are true.

```r
if(<condition1>) {

}

if(<condition2>) {

}
```


## `for` Loops

[Watch a video of this section](https://youtu.be/FbT1dGXCCxU)

For loops are pretty much the only looping construct that you will
need in R. While you may occasionally find a need for other types of
loops, in my experience doing data analysis, I've found very few
situations where a for loop wasn't sufficient. 

In R, for loops take an interator variable and assign it successive
values from a sequence or vector. For loops are most commonly used for
iterating over the elements of an object (list, vector, etc.)

```{r}
for(i in 1:10) {
        print(i)
}
```

This loop takes the `i` variable and in each iteration of the loop
gives it values 1, 2, 3, ..., 10, executes the code within the curly
braces, and then the loop exits.

The following three loops all have the same behavior.

```{r}
x <- c("a", "b", "c", "d")

for(i in 1:4) {
        ## Print out each element of 'x'
        print(x[i])  
}
```

The `seq_along()` function is commonly used in conjunction with for
loops in order to generate an integer sequence based on the length of
an object (in this case, the object `x`).

```{r}
## Generate a sequence based on length of 'x'
for(i in seq_along(x)) {   
        print(x[i])
}
```

It is not necessary to use an index-type variable. 

```{r}
for(letter in x) {
        print(letter)
}
```

For one line loops, the curly braces are not strictly necessary.

```{r}
for(i in 1:4) print(x[i])
```

However, I like to use curly braces even for one-line loops, because
that way if you decide to expand the loop to multiple lines, you won't
be burned because you forgot to add curly braces (and you *will* be
burned by this).


## Nested `for` loops

`for` loops can be nested inside of each other.

```r
x <- matrix(1:6, 2, 3)

for(i in seq_len(nrow(x))) {
        for(j in seq_len(ncol(x))) {
                print(x[i, j])
        }   
}
```

Nested loops are commonly needed for multidimensional or hierarchical
data structures (e.g. matrices, lists). Be careful with nesting
though. Nesting beyond 2 to 3 levels often makes it difficult to
read/understand the code. If you find yourself in need of a large
number of nested loops, you may want to break up the loops by using
functions (discussed later).


## `while` Loops

[Watch a video of this section](https://youtu.be/VqrS1Wghq1c)

While loops begin by testing a condition. If it is true, then they
execute the loop body. Once the loop body is executed, the condition
is tested again, and so forth, until the condition is false, after
which the loop exits.

```{r}
count <- 0
while(count < 10) {
        print(count)
        count <- count + 1
}
```

While loops can potentially result in infinite loops if not written
properly. Use with care!

Sometimes there will be more than one condition in the test.

```{r}
z <- 5
set.seed(1)

while(z >= 3 && z <= 10) {
        coin <- rbinom(1, 1, 0.5)
        
        if(coin == 1) {  ## random walk
                z <- z + 1
        } else {
                z <- z - 1
        } 
}
print(z)
```

Conditions are always evaluated from left to right. For example, in
the above code, if `z` were less than 3, the second test would not
have been evaluated.



## `repeat` Loops

[Watch a video of this section](https://youtu.be/SajmdYr30SY)

`repeat` initiates an infinite loop right from the start. These are
not commonly used in statistical or data analysis applications but
they do have their uses. The only way to exit a `repeat` loop is to
call `break`.

One possible paradigm might be in an iterative algorithm where you may
be searching for a solution and you don't want to stop until you're
close enough to the solution. In this kind of situation, you often
don't know in advance how many iterations it's going to take to get
"close enough" to the solution.

```{r,eval=FALSE,prompt=FALSE}
x0 <- 1
tol <- 1e-8

repeat {
        x1 <- computeEstimate()
        
        if(abs(x1 - x0) < tol) {  ## Close enough?
                break
        } else {
                x0 <- x1
        } 
}
```

Note that the above code will not run if the `computeEstimate()`
function is not defined (I just made it up for the purposes of this
demonstration).

The loop above is a bit dangerous because there's no guarantee it will
stop. You could get in a situation where the values of `x0` and `x1`
oscillate back and forth and never converge. Better to set a hard
limit on the number of iterations by using a `for` loop and then
report whether convergence was achieved or not.


## `next`, `break`

`next` is used to skip an iteration of a loop. 

```{r,eval=FALSE,prompt=FALSE}
for(i in 1:100) {
        if(i <= 20) {
                ## Skip the first 20 iterations
                next                 
        }
        ## Do something here
}
```

`break` is used to exit a loop immediately, regardless of what
iteration the loop may be on.

```{r,eval=FALSE,prompt=FALSE}
for(i in 1:100) {
      print(i)

      if(i > 20) {
              ## Stop loop after 20 iterations
              break  
      }		
}
```


## Summary

- Control structures like `if`, `while`, and `for` allow you to
  control the flow of an R program

- Infinite loops should generally be avoided, even if (you believe)
  they are theoretically correct.

- Control structures mentioned here are primarily useful for writing
  programs; for command-line interactive work, the "apply" functions
  are more useful.
